Death by Dropdown? Engineer Insightful Shiny Apps with Behavioral Science

### 2025-09-07 conference-materials/2025_posit-conf/draft.qmd
Shorten and refocus Posit Conf draft presentation
- Removed `custom.css` inclusion for simplicity and reduced styling overhead.
- Eliminated verbose narrative content in INTERPRET and NOTICE sections to fit tighter time constraints.
- Replaced text-heavy slides with concise, insight-focused messaging and behavioral science notes.
- Consolidated duplicate or overlapping slide content across NOTICE, ANTICIPATE, and STRUCTURE sections.
- Embedded code examples and dynamic illustrations to enhance clarity without lengthy explanations.

---
title: "Death by Dropdown? Engineer Insightful Shiny Apps with Behavioral Science"
author: "Jeremy Winget, PhD"
execute:
  echo: true
format:
  revealjs:
    theme: default
    css: custom.css
    slide-number: true
    transition: fade
---

@@

::: {.fragment .large}
**Or just get the urge to...**<br>
<div class="tenor-gif-embed" data-postid="17038619" data-share-method="host" data-aspect-ratio="1.77778" data-width="100%"><a href="https://tenor.com/view/ron-swanson-parks-and-recreation-parks-and-rec-computer-dumpster-gif-17038619">Ron Swanson Parks And Recreation GIF</a>from <a href="https://tenor.com/search/ron+swanson-gifs">Ron Swanson GIFs</a></div> <script type="text/javascript" async src="https://tenor.com/embed.js"></script>
:::

@@

**"Interfaces Need A Smart Vision"** - Remember: I-N-A-S-V

@@

::: {.large}
**Before starting the engine, know where you're going and why**
:::

## INTERPRET: Clarity Through Story

:::: {.columns}
::: {.column width="50%"}
### Key Questions
- **Who's driving?** Product managers
- **Where are we going?** "Are we on track this quarter?"
- **Why this route?** Tell a data story
:::

::: {.column width="50%"}
### The Data Story Arc
1. **Hook:** "Performance looks strong..."
2. **Context:** "New product launched"
3. **Tension:** "But churn is rising"
4. **Resolution:** "Focus on key segments"
:::
::::

::: {.notes}
Write the travel guide before packing the car
:::

## INTERPRET: With {bidux}

@@

# Define your destination and purpose
interpret_stage <- bid_interpret(
  central_question = "Are we on track this quarter?",
  data_story = list(
    hook = "Q3 performance exceeds targets by 15%",
    context = "Following our new product launch",
    tension = "However, churn increased 8% in enterprise segment",
    resolution = "Focus retention efforts on high-value accounts"
  ),
  user_personas = list(
    primary = "Product managers checking weekly KPIs",
    secondary = "Executives reviewing quarterly trends"
  )
)
```

## NOTICE: Check the Warning Lights ⚠️ {background-color="#4682b4"}

::: {.large}
**Spot the mechanical issues before leaving the driveway**
:::

## When the client asks for "just one more filter"

![](img/anakin.jpg)

::: {.fragment}
**Spoiler:** It's never just one more
:::

## NOTICE: Identify the Real Problems

![](img/notice-markup.png)

:::: {.columns}
::: {.column width="50%"}
### Warning Lights On

- **Choice overload:** 18 filters = paralysis
- **No clear path:** Where to click first?
- **Hidden costs:** Cognitive burden
:::

::: {.column width="50%"}
### Diagnostic Results

- 65% of sessions end without selection
- Average time to first insight: 4.3 minutes
- Most used filters: Only 3 of 18
:::
::::

## NOTICE: With {bidux}

@@

# Real telemetry data reveals hidden friction
notice_stage <- bid_notice(
  previous_stage = interpret_stage,
  problem = "Analysis paralysis from filter overload",
  evidence = list(
    quantitative = "65% abandonment rate; 4.3 min to first click",
    qualitative = "Users report feeling overwhelmed",
    telemetry = "Heatmap shows 80% ignore bottom filters"
  )
)
```

## ANTICIPATE: Pack for Weather Changes 🌦️ {background-color="#4682b4"}

::: {.large}
**Prepare for the cognitive detours ahead**
:::

## ANTICIPATE: Guard Against Misinterpretation

@@

:::: {.fragment .columns}
::: {.column width="50%"}
### Common Cognitive Hazards

- **Anchoring bias:** First number sticks
- **Framing effects:** Progress vs. gap
- **Confirmation bias:** Cherry-picking data
:::

::: {.column width="50%"}
### Mitigation Strategies

![](img/framing-toggle.png)

- Toggle: "75% to goal" ↔ "25% gap"
- Reference lines with context
- Uncertainty indicators
:::
::::

## ANTICIPATE: Telemetry-Driven Example

@@

anticipate_stage <- bid_anticipate(
  previous_stage = notice_stage,
  include_accessibility = TRUE
)
```

## STRUCTURE: Map the Scenic Route 🛣️ {background-color="#4682b4"}

::: {.large}
**Design the journey, not just the destination**
:::

## The Universal Truth of Dashboard Design

@@

::: {.fragment}
**Pro tip:** If everything's important, nothing is
:::

## STRUCTURE: The Path to Insight

::: {.large}
**Main highway first. Scenic detours later.**
:::

:::: {.columns}
::: {.column width="50%"}
### Highway Design

- **Smart defaults** answer 80% of questions
- **Progressive disclosure** for complexity
- **Visual hierarchy** guides the eye
:::

::: {.column width="50%"}
### Rest Stops

- Group related controls
- Breathing room between sections
- Clear signposts at decision points
:::
::::

## STRUCTURE: Before → After Transformation

:::: {.columns}
::: {.column width="50%"}
### Before: Lost in the Wilderness

- Filters scattered like roadside litter
- Competing charts = conflicting GPS voices
- No defaults = starting from nowhere
:::

::: {.column width="50%"}
### After: Scenic Highway

![](img/after-structured.png)

- **Hero metric** answers the hook first
- **Guided facets** for exploration
- **Smart defaults** = instant value
:::
::::

## STRUCTURE: With {bidux}

@@

structure_stage <- bid_structure(previous_stage = anticipate_stage)
```

## VALIDATE: Confirm Safe Arrival ✅ {background-color="#4682b4"}

::: {.large}
**End the journey with clarity and next steps**
:::

## VALIDATE: The Destination Check

:::: {.columns}
::: {.column width="50%"}
### Did We Arrive?

- Plain language summary
- Key insights highlighted
- Confidence indicators
:::

::: {.column width="50%"}
### What's Next?

![](img/validate-actions.png)

- 2-3 recommended actions
- One-click report export
- Share with team button
:::
::::

::: {.notes}
Like the GPS saying "You have arrived at your destination"
:::

## When Users Finally Find What They Need

![](https://media.giphy.com/media/OK27wINdQS5YQ/giphy.gif)

::: {.fragment}
**This is what we're optimizing for!**
:::

## VALIDATE: With {bidux}

```{r}
#| eval: false
validate_stage <- bid_validate(
  previous_stage = structure_stage,
  summary_style = "Executive Brief",
  include_telemetry = TRUE,
  include_empower_tools = TRUE
)

summary(validate_stage)
```

## The Complete Journey: Before → After

:::: {.columns}
::: {.column width="50%"}
### Before: Lost Driver 😰

- 18 filters, 6 tabs
- No clear path
- 65% abandonment
- 4.3 min to insight
:::

::: {.column width="50%"}
### After: Smooth Highway 🚗

![](img/after-complete.png)

- 3 smart filters
- Guided journey
- 85% completion
- 45 sec to insight
:::
::::

## Real Production Results

@@

# Actual telemetry from bidux-powered dashboards
bid_report(
  app_id = "quarterly_kpi_dashboard",
  period = "2024_Q4"
)
```

::: {.fragment}
### Impact Metrics 📊

| Metric | Before BID | After BID | Change |
|--------|-----------|-----------|---------|
| Time to First Insight | 4.3 min | 45 sec | **-82%** |
| Session Completion | 35% | 85% | **+143%** |
| User Satisfaction (NPS) | 23 | 67 | **+44 pts** |
| Misinterpretation Rate | 31% | 8% | **-74%** |
| Weekly Active Users | 127 | 412 | **+224%** |
:::

## Key Takeaways 🎯

::: {.incremental}
1. **Ground design in behavioral science** → Less user friction, more insights

2. **Structure is the work** → The path matters more than the features

3. **Telemetry reveals hidden problems** → What users do > what they say

4. **Tools implement the science** → reactable, echarts4r, plotly + {bidux} guidance

5. **{bidux} = behavioral scientist in your console** → Evidence-based decisions at each step
:::

## And Remember! 🧠 {.center background-color="#2c3e50"}

@@

:::: {.columns}
::: {.column width="60%"}
### The BID Journey Recap 🚗

- **I**nterpret → Plan your destination
- **N**otice → Check warning lights
- **A**nticipate → Pack for weather
- **S**tructure → Map the route
- **V**alidate → Confirm arrival
:::

::: {.column width="40%"}
::: {.fragment}
<div class="tenor-gif-embed" data-postid="5313946" data-share-method="host" data-aspect-ratio="1.3" data-width="100%"><a href="https://tenor.com/view/writethatdown-gif-5313946">Writethatdown GIF</a>from <a href="https://tenor.com/search/writethatdown-gifs">Writethatdown GIFs</a></div> <script type="text/javascript" async src="https://tenor.com/embed.js"></script>
:::
:::
::::

::: {.notes}
- for gif: "Write that down! Stick it to your monitor!"
:::

## Your Action Items 🚀 {background-color="#27ae60"}

::: {.r-fit-text}
**Stop death by dropdown!**
:::

### Three Things to Do This Week:

::: {.incremental}
1. **Download {bidux}** → `install.packages("bidux")`
2. **Try a some of the auto-suggestions** on a completed/WIP dashboard
3. **Connect with me** to share your before/after → [@jrwinget](https://bsky.app/profile/jrwinget.bsky.social)
:::

::: {.fragment}
**Bonus:** First 5 people to share their transformation get free 1:1 dashboard review! 🎁
:::

## Let's Connect! 🤝 {.center background-color="#2c3e50"}

![QR Code for my info](img/qr-winget.png)

:::: {.columns}
::: {.column width="33%"}
### 📦 Get {bidux}

`install.packages("bidux")`
:::

::: {.column width="33%"}
### 💬 Share Your Story

[@jrwinget](https://bsky.app/profile/jrwinget.bsky.social)
:::

::: {.column width="33%"}
### 📧 Questions?

![](https://media.giphy.com/media/3o7btZ1Gm7ZL25pLMs/giphy.gif)
:::
::::

::: {.notes}
- Remind about mnemonic: "Interfaces Need A Smart Vision"
- Feel free to reach out if you'd like to connect and chat further
- Thank organizers
- Quick reminder about the free dashboard reviews
:::

>>>>

---
title: "Death by Dropdown? Engineer Insightful Shiny Apps with Behavioral Science"
author: "Jeremy Winget, PhD"
execute:
  echo: true
format:
  revealjs:
    theme: default
    slide-number: true
    transition: fade
---

@@

::: {.fragment .large}
**Or just get the urge to...**

<div class="tenor-gif-embed" data-postid="17038619" data-share-method="host" data-aspect-ratio="1.77778" data-width="100%"><a href="https://tenor.com/view/ron-swanson-parks-and-recreation-parks-and-rec-computer-dumpster-gif-17038619">Ron Swanson Parks And Recreation GIF</a>from <a href="https://tenor.com/search/ron+swanson-gifs">Ron Swanson GIFs</a></div> <script type="text/javascript" async src="https://tenor.com/embed.js"></script>
:::

@@

**Remember - "Interfaces Need A Smart Vision"**

@@

@@

# Define your destination and purpose
interpret_stage <- bid_interpret(
  central_question = "Are we on track this quarter?",
  data_story = list(
    hook = "Q3 performance exceeds targets by 15%",
    context = "Following our new product launch",
    tension = "However, churn increased 8% in enterprise segment",
    resolution = "Focus retention efforts on high-value accounts"
  ),
  user_personas = list(
    primary = "Product managers checking weekly KPIs",
    secondary = "Executives reviewing quarterly trends"
  )
)

# Quick destination check - devs often skip this
bid_interpret(
  central_question = "Are we on track this quarter?",
  user_persona = "PM checking weekly KPIs"  # ← Most forget this
)
```

::: {.fragment}
**Key insight:** If you can't write the question in 10 words, users won't find answers in 100 filters
:::

::: {.notes}
- 1.5 minutes TOTAL for Interpret
- "You think you do this, but telemetry says otherwise"
- "Most jump straight to filters"
- Transition: "But here's what you're missing..."
:::

## NOTICE: The Warning Lights You're Ignoring ⚠️ {background-color="#e74c3c"}

![](img/telemetry-heatmap.png)

::: {.incremental}
- **Your assumption:** "They need all 18 filters"
- **Reality:** 82% use only 3 filters
- **Hidden cost:** 4.3 minutes to first click
:::

::: {.notes}
- START 3-minute NOTICE section
- "This is where behavioral science starts"
:::

## NOTICE: What Telemetry Reveals

@@

# Actual production telemetry
notice_results <- bid_notice(telemetry_data = TRUE)

summary(notice_results)
```

::: {.fragment}
**The "Just One More Filter" Death Spiral:**

![](img/anakin.jpg)
:::

::: {.notes}
- "Telemetry is your behavioral microscope"
- "Every hover tells a story of confusion"
- Transition: "Now let's talk about what happens in their heads"
:::

## NOTICE: Cognitive Load Measurement

:::: {.columns}
::: {.column width="50%"}
### You're Measuring:
- Click-through rates
- Time on page
- Completions
:::

::: {.column width="50%"}
### You Should Measure:
- **Decision fatigue points**
- **Paradox of choice moments**
- **Cognitive overflow indicators**
:::
::::

# update example based on actual bidux code
```

::: {.notes}
- "This is behavioral science in action"
- End NOTICE section (3 min total)
- Transition: "Once you spot problems, anticipate how users will cope"
:::

## ANTICIPATE: Pack for Cognitive Weather 🌦️ {background-color="#f39c12"}

::: {.large}
**Users don't see data. They see stories.**
:::

@@

::: {.fragment}
*And they'll write their own story if you don't guide them*
:::

::: {.notes}
- START 3-minute ANTICIPATE section
- "This is pure psychology"
- So we need to prepare for cognitive detours
:::

## ANTICIPATE: The Biases You're Not Defending Against

:::: {.columns}
::: {.column width="50%"}
### Anchoring Bias
```{r}
#| eval: false

# Bad: First number sticks
valueBox(value = -5, subtitle = "Change")

# Good: Context first
valueBox(value = "+15% YTD", subtitle = "Despite -5 this week")
```
:::

::: {.column width="50%"}
### Framing Effects
```{r}
#| eval: false

# bidux suggests toggle
bid_anticipate() |>
  add_framing_toggle(
    positive = "75% to goal",
    negative = "25% gap remaining"
  )
```
:::
::::

::: {.notes}
- "Small changes, massive perception shifts"
- "This is what behavioral scientists know that devs don't"
:::

## ANTICIPATE: Confirmation Bias Protection

![](img/confirmation-bias-toggle.png)

@@

# update with actual bidux code and set eval to TRUE
```

::: {.fragment}
**Result:** 74% reduction in misinterpretation
:::

::: {.notes}
- "Protect users from themselves"
- End ANTICIPATE (3 min total)
- Transition: "Now structure the journey"
:::

## STRUCTURE: The Scenic Route 🛣️ {background-color="#16a085"}

@@

::: {.fragment}
**If everything's important, nothing is.**
:::

::: {.notes}
- START 3.5-minute STRUCTURE section
- "This is where psychology meets code"
:::

## STRUCTURE: Progressive Disclosure That Works

:::: {.columns}
::: {.column width="50%"}
### Before: Chaos
![](img/before-overload.png)

- All options visible
- Competing for attention
- Analysis paralysis
:::

::: {.column width="50%"}
### After: Journey
![](img/after-structured.png)

- Start with one question
- Reveal complexity gradually
- Cognitive scaffolding
:::
::::

::: {.notes}
- "Structure IS the user experience"
- "Every pixel placement is a psychological choice"
:::

## STRUCTURE: The Behavioral Science Hierarchy

@@

# update with actual bidux code and set eval to TRUE
```

::: {.fragment}
### The F-Pattern Reading Path
![](img/f-pattern-heatmap.png)

*Place critical info where eyes naturally go*
:::

::: {.notes}
- "Eye tracking shows F-pattern on dashboards"
- "Top-left = highest value real estate"
:::

## STRUCTURE: Smart Defaults from Telemetry

@@

# update with actual bidux code and set eval to TRUE
```

::: {.fragment}
**Before:** User configures dashboard<br>
**After:** Dashboard anticipates user
:::

::: {.notes}
- End STRUCTURE (3.5 min total)
- Transition: "Quick validation check"
:::

## VALIDATE: Confirm Safe Arrival ✅ {background-color="#27ae60"}

# update with actual bidux code and set eval to TRUE
```

::: {.fragment}
![](img/validate-summary.png)

**The GPS moment:** "You've arrived at your insight"
:::

::: {.notes}
- 1.5 minutes TOTAL for Validate
- "Devs know summaries, but miss the psychology"
- "Plain language = cognitive relief"
- Transition: "Let's see the impact"
:::

## The Transformation: Proven Results 📊

:::: {.columns}
::: {.column width="50%"}
### Before BID
- 18 filters, 6 tabs
- 4.3 min to insight
- 65% abandonment
- 31% misinterpret data
:::

::: {.column width="50%"}
### After BID
- 3 smart filters
- 45 sec to insight (-82%)
- 85% completion (+143%)
- 8% misinterpret (-74%)
:::
::::

::: {.fragment}
### One Team's Story:
*"We thought users wanted more options. Telemetry showed they wanted more guidance. BID framework + {bidux} = 224% increase in weekly active users"*
:::

::: {.notes}
- 1.5 minutes for results
- "This isn't theory - it's production data"
- Transition: "Here's how to start"
:::

## Your Roadmap Forward 🚀 {.center background-color="#2c3e50"}

@@

### This Week's Challenge:

1. **Install {bidux}:** `install.packages("bidux")`
2. **Run telemetry on ONE dashboard:** `bid_notice(telemetry = TRUE)`
3. **Share your before/after:** [@jrwinget](https://bsky.app/profile/jrwinget.bsky.social)

::: {.fragment}
**🎁 First 5 transformations get free 1:1 dashboard review!**

![QR for contact](img/qr-combined.png)
:::

::: {.notes}
- 1 minute close
- "No more death by dropdown!"
- "Questions?"
- Thank organizers if time
:::

## Bonus Slide: The {bidux} Toolkit {.smaller}

# For those who want to dive deep
library(bidux)

# Full workflow in 5 lines
my_dashboard |>
  bid_interpret(question = "What's our KPI status?") |>
  bid_notice(telemetry = TRUE) |>
  bid_anticipate(guard_biases = TRUE) |>
  bid_structure(progressive = TRUE) |>
  bid_validate(empower_users = TRUE)

# Generate full behavioral report
bid_report(my_dashboard, output = "behavioral_insights.html")
```

::: {.notes}
- Only if extra time in Q&A
- "Full framework in your console"
:::

## Bonus: Quick Reference Card

| Stage | Problem | Solution | {bidux} Function |
|-------|---------|----------|------------------|
| **I**nterpret | No clear question | Define destination | `bid_interpret()` |
| **N**otice | Hidden friction | Telemetry reveals | `bid_notice()` |
| **A**nticipate | Cognitive biases | Guard rails | `bid_anticipate()` |
| **S**tructure | Chaos overwhelms | Progressive disclosure | `bid_structure()` |
| **V**alidate | Unclear outcomes | Plain language | `bid_validate()` |

::: {.notes}
- Only show if someone asks for summary
- "Email me for full guide"
:::
